---
title: "Check the project"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: true
  message: true
  warning: true
params:
  strict: false
---

This Quarto document runs lightweight local checks for the template structure,
required software packages, generated outputs, privacy-related ignore rules, and
obvious local path references.

The checks are intentionally simple and editable. The default example uses R,
but projects that add Python, Julia, or other languages can add extra code
chunks below without changing the project layout.

Run from the project root with:

```sh
quarto render code/check-project.qmd
```

Use strict mode to treat warnings as failures:

```sh
quarto render code/check-project.qmd -P strict:true
```

```{r}
#| label: setup
find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, mustWork = TRUE)

  repeat {
    if (
      file.exists(file.path(current, "README.md")) &&
        file.exists(file.path(current, "project-metadata.yml"))
    ) {
      return(current)
    }

    parent <- dirname(current)
    if (identical(parent, current)) {
      stop("Could not find the project root.", call. = FALSE)
    }
    current <- parent
  }
}

project_root <- find_project_root()
knitr::opts_knit$set(root.dir = project_root)

is_true <- function(value) {
  identical(value, TRUE) || identical(tolower(as.character(value)), "true")
}

strict <- is_true(params$strict)
failures <- character()
check_warnings <- character()

add_failure <- function(message) {
  failures <<- c(failures, message)
}

add_warning <- function(message) {
  check_warnings <<- c(check_warnings, message)
}

check_files_exist <- function(paths, label) {
  missing <- paths[!file.exists(paths)]
  if (length(missing) > 0) {
    add_failure(paste0("Missing ", label, ": ", paste(missing, collapse = ", ")))
  }
}

check_dirs_exist <- function(paths, label) {
  missing <- paths[!dir.exists(paths)]
  if (length(missing) > 0) {
    add_failure(paste0("Missing ", label, ": ", paste(missing, collapse = ", ")))
  }
}
```

## Required Structure

```{r}
#| label: required-structure
required_dirs <- c(
  "assets",
  "code",
  "code/analysis",
  "code/exploration",
  "code/processing",
  "code/utilities",
  "data",
  "data/raw-data",
  "data/processed-data",
  "data/private-data",
  "data/large-files",
  "products",
  "results",
  "results/figures",
  "results/tables",
  "ai"
)

required_files <- c(
  "README.md",
  "AGENTS.md",
  "project-metadata.yml",
  "data/data-use-notes.md",
  "ai/README.md",
  "ai/ai-use-log.md",
  "ai/review-checklist.md",
  "code/run-all.qmd",
  "code/check-project.qmd",
  "code/workflow.md",
  "code/processing/processingcode.R",
  "code/exploration/edacode.R",
  "code/analysis/statistical-analysis.R",
  "code/utilities/README.md",
  "products/report/report.qmd",
  "products/manuscript/Manuscript.qmd",
  "products/presentation/presentation.qmd"
)

check_dirs_exist(required_dirs, "required folder(s)")
check_files_exist(required_files, "required file(s)")

structure_table <- data.frame(
  path = c(required_dirs, required_files),
  type = c(rep("folder", length(required_dirs)), rep("file", length(required_files))),
  present = c(dir.exists(required_dirs), file.exists(required_files))
)

knitr::kable(structure_table)
```

## Required R Packages

The example workflow uses these R packages. Do not silently install missing
packages; report them so the project owner or student can install them
deliberately.

```{r}
#| label: package-checks
required_packages <- c(
  "broom",
  "dplyr",
  "ggplot2",
  "here",
  "knitr",
  "readxl",
  "skimr",
  "tidyr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  add_failure(
    paste0(
      "Missing required R package(s): ",
      paste(missing_packages, collapse = ", "),
      ". Install them before running the example workflow."
    )
  )
}

package_table <- data.frame(
  package = required_packages,
  available = !(required_packages %in% missing_packages)
)

knitr::kable(package_table)
```

## Generated Outputs

```{r}
#| label: output-checks
example_outputs <- c(
  "data/processed-data/processeddata.rds",
  "results/tables/summarytable.rds",
  "results/tables/summarytable.csv",
  "results/tables/resulttable1.rds",
  "results/tables/resulttable1.csv",
  "results/tables/resulttable2.rds",
  "results/tables/resulttable2.csv",
  "results/figures/height-distribution.png",
  "results/figures/weight-distribution.png",
  "results/figures/height-weight.png",
  "results/figures/height-weight-stratified.png"
)

missing_outputs <- example_outputs[!file.exists(example_outputs)]
if (length(missing_outputs) > 0) {
  add_warning(
    paste0(
      "Example generated output(s) are missing: ",
      paste(missing_outputs, collapse = ", "),
      ". Run quarto render code/run-all.qmd or adjust this check for your project."
    )
  )
}

output_table <- data.frame(
  output = example_outputs,
  present = file.exists(example_outputs)
)

knitr::kable(output_table)
```

## Privacy and Local Files

```{r}
#| label: privacy-checks
if (file.exists(".gitignore")) {
  gitignore <- readLines(".gitignore", warn = FALSE)
  expected_ignore_rules <- c(
    ".Renviron",
    ".env",
    "data/private-data/*",
    "data/large-files/*"
  )
  missing_rules <- expected_ignore_rules[
    !vapply(
      expected_ignore_rules,
      function(rule) any(trimws(gitignore) == rule),
      logical(1)
    )
  ]

  if (length(missing_rules) > 0) {
    add_warning(
      paste0(
        ".gitignore is missing recommended privacy/local-data rule(s): ",
        paste(missing_rules, collapse = ", ")
      )
    )
  }
} else {
  add_failure("Missing .gitignore.")
}

local_secret_files <- c(".Renviron", ".env")
present_secret_files <- local_secret_files[file.exists(local_secret_files)]
if (length(present_secret_files) > 0) {
  add_warning(
    paste0(
      "Local secret/environment file(s) are present: ",
      paste(present_secret_files, collapse = ", "),
      ". Confirm they are ignored and do not commit them."
    )
  )
}
```

## Local Path Scan

```{r}
#| label: local-path-scan
root_text_files <- c("README.md", "AGENTS.md", "project-metadata.yml")
root_text_files <- root_text_files[file.exists(root_text_files)]

text_files <- c(
  root_text_files,
  list.files(
  c("code", "products", "ai", "data", "results"),
  pattern = "\\.(R|qmd|md|py|jl|yml|yaml)$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
  )
)

text_files <- setdiff(
  text_files,
  c(file.path("code", "check-project.qmd"))
)

absolute_path_pattern <- "[A-Za-z]:\\\\|/Users/|/home/"
path_hits <- character()

for (path in text_files) {
  lines <- readLines(path, warn = FALSE)
  hit_index <- grep(absolute_path_pattern, lines)

  if (length(hit_index) > 0) {
    hit_lines <- lines[hit_index]
    explanatory <- grepl(
      "absolute path|you never want|C:\\\\yourname|C:\\\\ahandel",
      hit_lines,
      ignore.case = TRUE
    )
    hit_index <- hit_index[!explanatory]
  }

  if (length(hit_index) > 0) {
    path_hits <- c(path_hits, paste0(path, ":", hit_index))
  }
}

if (length(path_hits) > 0) {
  add_warning(
    paste0(
      "Potential absolute local path reference(s): ",
      paste(path_hits, collapse = ", ")
    )
  )
}

if (length(path_hits) == 0) {
  message("No unexpected absolute local path references found.")
}
```

## Optional Multi-Language Checks

If the project adds other languages, add small checks here. Keep them explicit
and local rather than silently installing software or packages. The examples
below are plain fenced code blocks so the base template does not require Python,
Julia, or extra Quarto engines. Once a project uses those languages, convert the
relevant example into an executable Quarto chunk.

```python
# Example Python package check:
import importlib.util

missing = [
    package for package in ["pandas", "numpy"]
    if importlib.util.find_spec(package) is None
]

if missing:
    raise SystemExit(f"Missing Python package(s): {', '.join(missing)}")
```

```julia
# Example Julia environment check:
import Pkg
Pkg.status()
```

## Summary

```{r}
#| label: summary
if (strict && length(check_warnings) > 0) {
  failures <- c(failures, paste0("Strict-mode warning: ", check_warnings))
}

cat("Project check\n")
cat("Working directory: ", project_root, "\n", sep = "")

if (length(check_warnings) > 0) {
  cat("\nWarnings:\n")
  for (warning in check_warnings) {
    cat("- ", warning, "\n", sep = "")
  }
}

if (length(failures) > 0) {
  cat("\nFailures:\n")
  for (failure in failures) {
    cat("- ", failure, "\n", sep = "")
  }
  stop(
    paste(c("Project checks failed:", paste0("- ", failures)), collapse = "\n"),
    call. = FALSE
  )
}

cat("\nOK: project checks passed.\n")
```
