---
title: "Run the project workflow"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: true
  message: true
  warning: true
params:
  render_products: false
  render_supplement: false
  skip_checks: false
  strict_checks: false
---

This Quarto document is the main reproducibility entry point for the example
project. It runs the workflow steps in order and can optionally render the
project products.

The default example uses R, but the workflow is intentionally command-oriented.
Projects can add Python, Julia, shell, or other steps by adding them to the
workflow list below.

Run the default workflow from the project root with:

```sh
quarto render code/run-all.qmd
```

To also render the report, manuscript, and presentation:

```sh
quarto render code/run-all.qmd -P render_products:true
```

To include the supplementary material, which may require a TeX installation:

```sh
quarto render code/run-all.qmd -P render_products:true -P render_supplement:true
```

```{r}
#| label: setup
find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, mustWork = TRUE)

  repeat {
    if (
      file.exists(file.path(current, "README.md")) &&
        file.exists(file.path(current, "project-metadata.yml"))
    ) {
      return(current)
    }

    parent <- dirname(current)
    if (identical(parent, current)) {
      stop("Could not find the project root.", call. = FALSE)
    }
    current <- parent
  }
}

project_root <- find_project_root()
knitr::opts_knit$set(root.dir = project_root)

is_true <- function(value) {
  identical(value, TRUE) || identical(tolower(as.character(value)), "true")
}

message("Project root: ", project_root)
```

## Workflow Steps

Each step should declare its role, language, inputs, and outputs. Keeping this
table current is more important than the exact folder layout, especially once a
project uses more than one programming language.

```{r}
#| label: workflow-table
workflow_steps <- list(
  list(
    stage = "processing",
    description = "Clean the example raw data.",
    language = "R",
    runner = "r-source",
    path = "code/processing/processingcode.R",
    command = "source",
    args = "code/processing/processingcode.R",
    inputs = "data/raw-data/exampledata.xlsx",
    outputs = "data/processed-data/processeddata.rds"
  ),
  list(
    stage = "exploration",
    description = "Create exploratory tables and figures.",
    language = "R",
    runner = "r-source",
    path = "code/exploration/edacode.R",
    command = "source",
    args = "code/exploration/edacode.R",
    inputs = "data/processed-data/processeddata.rds",
    outputs = "results/tables/summarytable.*, results/figures/*.png"
  ),
  list(
    stage = "analysis",
    description = "Fit the example statistical models.",
    language = "R",
    runner = "r-source",
    path = "code/analysis/statistical-analysis.R",
    command = "source",
    args = "code/analysis/statistical-analysis.R",
    inputs = "data/processed-data/processeddata.rds",
    outputs = "results/tables/resulttable*.rds, results/tables/resulttable*.csv"
  )
)

workflow_table <- data.frame(
  stage = vapply(workflow_steps, `[[`, character(1), "stage"),
  language = vapply(workflow_steps, `[[`, character(1), "language"),
  command = vapply(
    workflow_steps,
    function(step) paste(c(step$command, step$args), collapse = " "),
    character(1)
  ),
  inputs = vapply(workflow_steps, `[[`, character(1), "inputs"),
  outputs = vapply(workflow_steps, `[[`, character(1), "outputs")
)

knitr::kable(workflow_table)
```

## Run Analysis Code

```{r}
#| label: run-workflow
run_r_script <- function(path) {
  if (!file.exists(path)) {
    stop("Cannot find R script: ", path, call. = FALSE)
  }

  message("\nRunning ", path)
  plot_file <- tempfile(fileext = ".pdf")
  grDevices::pdf(file = plot_file)

  tryCatch(
    sys.source(path, envir = new.env(parent = globalenv())),
    finally = {
      if (grDevices::dev.cur() > 1L) {
        grDevices::dev.off()
      }
      unlink(plot_file)
    }
  )
}

run_command <- function(command, args = character()) {
  executable <- Sys.which(command)
  if (!nzchar(executable) && file.exists(command)) {
    executable <- command
  }
  if (!nzchar(executable)) {
    stop("Cannot find command on PATH: ", command, call. = FALSE)
  }

  status <- system2(executable, args = args)
  if (!identical(status, 0L)) {
    stop("Command failed: ", command, " ", paste(args, collapse = " "), call. = FALSE)
  }
}

run_step <- function(step) {
  if (identical(step$runner, "r-source")) {
    run_r_script(step$path)
  } else if (identical(step$runner, "command")) {
    run_command(step$command, step$args)
  } else {
    stop("Unknown runner for workflow step: ", step$stage, call. = FALSE)
  }
}

for (step in workflow_steps) {
  run_step(step)
}
```

## Optional Product Rendering

```{r}
#| label: render-products
render_qmd <- function(path) {
  if (!file.exists(path)) {
    stop("Cannot find Quarto file: ", path, call. = FALSE)
  }

  quarto <- Sys.which("quarto")
  if (!nzchar(quarto)) {
    stop("Quarto was not found on PATH.", call. = FALSE)
  }

  message("\nRendering ", path)
  status <- system2(quarto, c("render", path))

  if (!identical(status, 0L)) {
    stop("Quarto render failed for ", path, call. = FALSE)
  }
}

if (is_true(params$render_products)) {
  render_qmd("products/report/report.qmd")
  render_qmd("products/manuscript/Manuscript.qmd")
  render_qmd("products/presentation/presentation.qmd")
} else {
  message("Product rendering skipped. Set render_products:true to enable it.")
}

if (is_true(params$render_supplement)) {
  render_qmd("products/manuscript/supplement/Supplementary-Material.qmd")
}
```

## Optional Checks

```{r}
#| label: run-checks
if (is_true(params$skip_checks)) {
  message("Project checks skipped.")
} else {
  quarto <- Sys.which("quarto")
  if (!nzchar(quarto)) {
    stop("Quarto was not found on PATH.", call. = FALSE)
  }

  check_args <- c("render", "code/check-project.qmd")
  if (is_true(params$strict_checks)) {
    check_args <- c(check_args, "-P", "strict:true")
  }

  message("\nRunning project checks")
  status <- system2(quarto, check_args)

  if (!identical(status, 0L)) {
    stop("Project checks failed.", call. = FALSE)
  }
}

message("\nWorkflow complete.")
```

## Adding Other Languages

Add other-language steps to `workflow_steps` with `runner = "command"`. For
example, a Python script could be listed as `command = "python"` and
`args = c("code/analysis/model.py")`. A Julia script could use
`command = "julia"` and `args = c("code/analysis/simulation.jl")`.

The corresponding code can still live in the workflow-stage folder where it
belongs. A project only needs language-specific subfolders if it becomes large
enough that separate packages, modules, or environments are easier to manage.
