---
title: "An example cleaning script"
author: "NAME"
date: "2026-01-01"
format:
  html:
    embed-resources: true
---


# Processing script

This contains the same code and comments/information as `processing-code.r`.

This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.

See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.


# Setup

Load needed packages. We first check that they are all available and stop with
an actionable message if one is missing. We only check, never install, since
installing packages from inside an analysis script would silently change the
environment the results were produced in.

```{r}
required_packages <- c("readxl", "dplyr", "tidyr", "skimr", "here")
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Required package '", pkg, "' is not installed.\n",
         "Install it with install.packages(\"", pkg, "\") and rerun this script.")
  }
}

library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data
library(here) #to set paths
```


# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

We stop early if the raw data is not where we expect it, or if it does not have
the columns the cleaning steps below rely on. A clear message here is much
easier to act on than an error from deep inside a later step.

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","example-data.xlsx")

if (!file.exists(data_location)) {
  stop("Raw data file not found at:\n  ", data_location,
       "\nCheck that you are working from the project root folder.")
}
if (!"Codebook" %in% readxl::excel_sheets(data_location)) {
  stop("The raw data file has no 'Codebook' sheet. Found: ",
       paste(readxl::excel_sheets(data_location), collapse = ", "))
}

rawdata <- readxl::read_excel(data_location)

required_columns <- c("Height", "Weight", "Gender")
if (!all(required_columns %in% names(rawdata))) {
  stop("Raw data is missing required column(s): ",
       paste(setdiff(required_columns, names(rawdata)), collapse = ", "))
}
```


# Check data

First we can look at the codebook

```{r}
codebook <- readxl::read_excel(data_location, sheet ="Codebook")
print(codebook)
```


Several ways of looking at the data

```{r}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```



# Cleaning

By inspecting the data as done above, we find some problems that need addressing:

First, there is an entry for height which says "sixty" instead of a number.
Does that mean it should be a numeric 60? It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
Since we don't know how to fix this, we might decide to remove the person. This "sixty" entry also turned all Height entries into characters instead of numeric. That conversion to character also means that our summary function isn't very meaningful. So let's fix that first.

```{r}
d1 <- rawdata %>% dplyr::filter( Height != "sixty" ) %>%
                  dplyr::mutate(Height = as.numeric(Height))
skimr::skim(d1)
hist(d1$Height)
```

Now we see that there is one person with a height of 6. That could be a typo, or someone mistakenly entered their height in feet. We assume that we know it was a mistaken entry as feet and convert to centimeters. Depending on your knowledge of the data, you might instead want to remove this person.

```{r}
d2 <- d1 %>% dplyr::mutate( Height = replace(Height, Height == 6, round(6*30.48,0)) )
skimr::skim(d2)
```

Height values seem ok now.

Now let's look at the `Weight` variable. There is a person with weight of 7000, which is impossible, and one person with missing weight.
To be able to analyze the data, we'll remove those individuals as well.

```{r}
d3 <- d2 %>%  dplyr::filter(Weight != 7000) %>% tidyr::drop_na()
skimr::skim(d3)
```

Now checking the `Gender` variable. Gender should be a categorical/factor variable but is loaded as character. We can fix that with simple base R code to mix things up.

```{r}
d3$Gender <- as.factor(d3$Gender)
skimr::skim(d3)
```


Now we see that there is another NA, but it's not `NA` from R, instead it was loaded as character and is now considered as a category.
There is also an individual coded as "N", which the codebook does not allow. This could be a mistyped M or a mistyped NA. If we had a good guess, we could adjust it; since we don't, we remove both the NA and the N individual here. Since this keeps empty categories for Gender, I'm also using droplevels() to get rid of them.

```{r}
d4 <- d3 %>% dplyr::filter( !(Gender %in% c("NA","N")) ) %>% droplevels()
skimr::skim(d4)
```


All done, data is clean now.

Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

```{r}
processeddata <- d4
```

Before saving, we check that cleaning removed exactly the individuals we expect
it to. This is the most important check here. Every step above removes rows, and
dropping the wrong rows is an easy mistake to make and a hard one to notice. We
know which 5 individuals should go: one with Height "sixty", one with Weight
7000, one with missing Weight, one with Gender "NA", and one with Gender "N".
If that number ever changes, the data or the cleaning logic changed, and we want
to stop and look rather than analyze a different sample by accident.

```{r}
expected_removed <- 5
actually_removed <- nrow(rawdata) - nrow(processeddata)
if (actually_removed != expected_removed) {
  stop("Cleaning removed ", actually_removed, " rows, but ", expected_removed,
       " were expected.\n",
       "Review the raw data and the cleaning steps before using these results.")
}
```


# Save data

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/RData files.
This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
save_data_location <- here::here("data","processed-data","processed-data.rds")
saveRDS(processeddata, file = save_data_location)
```



# Notes

Removing anyone who had "faulty" or missing data is one approach. It's often not the best. based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

