---
title: "An example exploratory analysis script with code pulled in"
author: "NAME"
date: "2026-01-01"
format:
  html:
    embed-resources: true
---



This Quarto file loads the cleaned data and does some exploring.

This is essentially the same as `eda.qmd`, but now the code is not inside this file. Instead, it is pulled in from the R script `eda-code.r` using the code chunk labels.

It is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files. Also often, an R script with enough comments also works well, and one doesn't need a Quarto file.

As in `eda.qmd`, the exploration code only computes summary results and saves them as an RDS file in `results/output/`. The polished tables and figures are produced by the scripts in `code/figures-tables/` (`make-tables.r` and `make-figures.r`), which read the saved results and processed data.

As part of the exploratory analysis, you should produce summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others.

Start by exploring one variable at a time. Then continue with the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.



# Setup
Load the chunk.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('eda-code.r')
```

Load the packages.
```{r,packages, echo=FALSE,message=FALSE}
```


Load the data.

```{r,loaddata}
```





# Data exploration through summary results

Showing a bit of code to compute a summary of the data and save the result.

```{r,table1}
```

We are saving only the *computed result* to `results/output/`. The finalized summary table (RDS plus a CSV companion) is produced later by `code/figures-tables/make-tables.r`, which reads this file. The exploratory figures (histograms and scatterplots) are produced by `code/figures-tables/make-figures.r`, which reads the processed data directly.


# Notes

For your own explorations, the summary results you compute can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures and tables you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible, which is why they are generated by the dedicated scripts in `code/figures-tables/`.
