---
title: "An example exploratory analysis script"
date: "2024-02-07"
format:
  html:
    embed-resources: true
---



This Quarto file loads the cleaned data and does some exploring.

This version has the code inside the Quarto file itself.

Also note that while cleaning and exploring is split here, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

This example separates *computing* results from *presenting* them. The exploration code here only computes summary results and saves them as an RDS file in `results/output/`. The polished tables and figures are then produced by the scripts in `code/figures-tables/` (`make-tables.r` and `make-figures.r`), which read the saved results and processed data. This keeps a single, clear place where every figure and table is generated.

As part of the exploratory analysis, you should produce summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others.

Start by exploring one variable at a time. Then continue with the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(skimr) #for a quick data summary
```


Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processed-data.rds")
#load data
mydata <- readRDS(data_location)
```





# Data exploration through summary results

Showing a bit of code to compute a summary of the data and save the result.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save the computed summary result (not a final table) to results/output
summary_result_file = here::here("results", "output", "summary-table.rds")
saveRDS(summary_df, file = summary_result_file)
```

We are saving only the *computed result* to `results/output/`. The finalized summary table (RDS plus a CSV companion) is produced later by `code/figures-tables/make-tables.r`, which reads this file. The exploratory figures (histograms and scatterplots) are produced by `code/figures-tables/make-figures.r`, which reads the processed data directly. 

This is a bit too much for a small project like this, but for larger projects, where code might take long to run, it is a good idea to split any kind of modeling/analysis from the generation of figures and tables. Keeping all figure and table generation in one folder makes it easy to see and rerun everything that turns results into presentation-ready outputs.


# Notes

For your own explorations, the summary results you compute can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures and tables you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible, which is why they are generated by the dedicated scripts in `code/figures-tables/`.
